---
title: "Getting Started with shapCopula"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with shapCopula}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 7, fig.height = 4
)
set.seed(42)
```

## Overview

`shapCopula` computes **Conditional SAGE** (Shapley Additive Global
importancE) values with asymptotically valid confidence intervals.
The key idea is to estimate the conditional value function
$$v(S;\, x) = \mathbb{E}[f(X) \mid X_S = x_S]$$
using a **Gaussian copula with empirical marginals**, then apply a
**one-step debiased estimator with K-fold cross-fitting** to obtain
root-n consistent, asymptotically normal SAGE estimates.

The companion paper is:

> Gnasso, A. (2026). *Inference for Conditional Shapley Values via Vine Copulas*. Manuscript under review.

## Installation

```r
# From GitHub (recommended)
remotes::install_github("agostinognasso/shapCopula")
```

## Minimal example

We construct a simple data-generating process where $X_1$ and $X_2$ are
correlated ($\rho = 0.6$) and $f$ depends linearly on $X_1$, $X_2$, and
$X_3$, with $X_4$ and $X_5$ irrelevant.

```{r dgp}
library(shapCopula)

n <- 500; p <- 5
Sigma <- diag(p)
Sigma[1, 2] <- Sigma[2, 1] <- 0.6   # X1 and X2 are correlated
L <- chol(Sigma)
X <- matrix(rnorm(n * p), n, p) %*% L
colnames(X) <- paste0("X", seq_len(p))

# A black-box predictor: truth is 2*X1 + X2 + 0.5*X3
true_coef <- c(2, 1, 0.5, 0, 0)
f_lin <- function(Xm) as.numeric(Xm %*% true_coef)
```

### Running the estimator

The main function is `sage_cv_gcop()`. Set `K = 3` folds,
`M = 30` permutations, `B = 20` conditional samples (default):

```{r run, cache = TRUE}
result <- sage_cv_gcop(f_lin, X, K = 3, M = 30, B = 20, seed = 1)
print(result, digits = 3)
```

Each row reports:

| Column | Meaning |
|--------|---------|
| `psi_hat` | SAGE estimate $\hat\Psi_j$ |
| `se` | Asymptotic standard error |
| `ci_lo`, `ci_hi` | 95% Wald confidence interval |
| `z` | Wald statistic for $H_0: \Psi_j = 0$ |
| `pvalue` | Two-sided p-value |

### Visualising the results

```{r plot, fig.alt = "SAGE bar chart with 95% CIs"}
ord <- order(result$psi_hat)
par(mar = c(4, 4, 2, 1))
barplot(
  result$psi_hat[ord],
  names.arg = result$feature[ord],
  horiz = TRUE, las = 1,
  xlab = "Conditional SAGE value",
  main = "Feature importance with 95% CIs"
)
# Add confidence intervals
y_pos <- barplot(result$psi_hat[ord], plot = FALSE)
arrows(result$ci_lo[ord], y_pos, result$ci_hi[ord], y_pos,
       angle = 90, code = 3, length = 0.05)
```

### Comparison with the marginal baseline

`sage_cv_marginal()` estimates the interventional SAGE (Janzing et al. 2020),
which ignores conditional dependence. Under correlation, it can
misattribute importance:

```{r marginal, cache = TRUE}
result_marg <- sage_cv_marginal(f_lin, X, K = 3, M = 30, B = 20, seed = 1)
comparison <- data.frame(
  feature    = result$feature,
  conditional = round(result$psi_hat, 3),
  marginal    = round(result_marg$psi_hat, 3)
)
print(comparison)
```

## Choosing the sampler

| Function | Sampler | When to use |
|----------|---------|-------------|
| `sage_cv_gcop()` | Gaussian copula | Default — fast, works for $p \lesssim 15$ |
| `sage_cv()` | Nonparametric vine copula | Stronger non-Gaussian dependence, $p \lesssim 10$ |
| `sage_cv_marginal()` | Marginal bootstrap | Interventional baseline / competitor |

## Tuning parameters

- **`K`** (folds): 3 is the standard choice; use 5 for small $n$.
- **`M`** (permutations): 30 gives stable estimates; increase to 50–100 for
  publication-quality results. Runtime scales linearly in $M$.
- **`B`** (conditional samples): 20 is default; the U-statistic debiasing
  uses two sub-batches of $B/2$, so the effective sample size is $B/2$.
  Increasing $B$ reduces Monte Carlo variance but increases runtime linearly.

Sensitivity of coverage and power to these choices is studied in Appendix C
of the companion paper.

## Reproducibility

Set the `seed` argument to any integer for deterministic results:

```r
sage_cv_gcop(f_lin, X, K = 3, M = 30, B = 20, seed = 2026)
```

## Session info

```{r session}
sessionInfo()
```
